---
title: "Lassen FIA Processing"
author: "Alexis Bernal"
date: "February 28, 2020"
output: word_document
---

#Loading libraries necessary to process FIA data

```{r}
library(raster) #Work with spatial
library(sf) #Work with spatial files
library(rgdal) #Work with spatial files
library(here) #Call on files from Collins_Pine-master R project
library(ggthemes) #manipulate aesthetics of ggplot
library(tidyverse) #Data wrangling and manipulation
```

#Section I - Working with spatial data to subset FIA plots

Creating a reproducible process of importing spatial data into R and selecting plots from FIA data to compare to Historical QQ data from Collins Pine. When looking for the appropriate "boundary" to constrain our sampling of FIA data, our choices were limited to Lassen/Plumas National Forest and multiple county boundaries (QQ plots located in Tehama and Plumas County... also added Lassen). Although we could look at NF or county boundaries separateley, I chose to inculde both to see if we could get a robust sample size for both public and private land for comparisons.  Parts of this section are pretty time-consuming processes (section distinguished) because the DEMs are such a huge dataset, so I exported the final clipped DEM to call on for future uses. #ABCHECKME I think this section needs updating (not urgent)

```{r}
#Loading DEM used by JTS for spatial extractions (originated from LANDFIRE)
#Currenlty two DEM files in Lassen_GIS_JTS; this one is from "Aligned"
#Not sure of the difference between Aligned and Aligned2
#Transforming raster to projection EPSG 4269 (similar to EDNF analysis)
DEM <- raster(here("large_files/Lassen_GIS_JTS/Aligned", "dem_aligned.tif")) %>%
  projectRaster(., crs = CRS("+init=EPSG:4269"))

#Looking at FIA spreadsheets, whole CN number is not visible.
#Setting up options in R to view all digits (max is 22).
options(digits = 22)

#Loading FIA plot data (USFS FIA)
FIA <- read.csv(here("large_files/CA_FIA", "CA_PLOTSNAP.csv"))

#Filtering FIA for inventory year (INVYR) > 2010 (measure in 10-yr cycles).
#Plots within these years have the same plot design protocol (DESIGNCD = 1).
#Plots within these years are also the most current (some plots are remeasured).
#Creating new ID's based on coordinates to identify remeasured plots.
#PREV_PLT_CN does indicate that some of these plots are re-measurements, but...
#Previous measurements taken before 2011 (so not "double counting" plots in this dataset).
#Filtered for most recent data (max INVYR), just in case there are duplicates.
#Technically 4-subplots/plot - so looks like duplicate entries for each year, but they're not.
#Removing duplicate rows for now (just need to subset plot locations).
#Extracting relevant columns for analysis.
#Adding coordinate columns to retain in sf dataframe.
#Converting FIA data to simple feature.
#Adding CRS EPSG 4269
FIA.sf <- FIA %>%
  filter(INVYR > 2010) %>%
  mutate(new.ID = group_indices(., LON, LAT)) %>%
  group_by(new.ID) %>%
  filter(INVYR == max(INVYR)) %>%
  ungroup() %>%
  mutate(duplicates = duplicated(new.ID)) %>%
  filter(duplicates == "FALSE") %>%
  dplyr::select(CN, PREV_PLT_CN, INVYR, PLOT, LON, LAT, ELEV, DESIGNCD) %>%
  mutate(x.coord = LON,
         y.coord = LAT) %>%
  st_as_sf(., coords = c("LON", "LAT")) %>%
  st_set_crs(., st_crs("EPSG:4269"))

#Creating a boundary for DEM to clip FIA points to (polygon)
DEM.boundary <- st_as_sfc(st_bbox(DEM))

#Clipping FIA points to DEM boundary
FIA.sf.clip <- st_intersection(FIA.sf, DEM.boundary)

#Checking alignment of layers
#Some points outside of actual raster data, probably due to projection?
#Those will be removed anyways based on guidelines for elevation and fuzzing
plot(DEM)
plot(FIA.sf.clip["geometry"], col = "red", add = TRUE)
plot(DEM.boundary, add = TRUE)

#Start with 274 plots (FIA plots within boundary).
#Extracting DEM values to FIA plots.
#Converting elevation of FIA plots (originally in feet) to meters.
#Selecting FIA plots within the elevation range of the QQ data (1113m - 1939m; removed 70 plots).
#Selecting plots within "fuzzing" standards (150 m max discrepancy; removed 22 plots).
#End with 182 plots.
elevation <- FIA.sf.clip %>%
  mutate(DEM.value = raster::extract(DEM,.),
         ELEV_met= as.double(ELEV/3.281),
         Elev.discrepancy = abs(ELEV_met - DEM.value)) %>%
  filter(ELEV_met > 1112 &
           ELEV_met < 1923) %>%
  filter(Elev.discrepancy < 151)
```

#Section II - Extracting plot data for plots with only 1 condition

First, created functions to estimate density (TPHA) and live BA (sq m/ha). Checking what the sample size would be given several constraints: only 1 condition (no split conditions within the plot), live BA > 9 m^2/ha, no fire damage, and no anomalous forest type. Our sample size started with 203 plots, after accounting for elevation discrepancies from fuzzing and being within the elevation range of the historical data. Found 81 plots had more than one condition and an additional 31 plots had fire associated with them (disturbance codes 30-32). Also found 1 additional plots with FORTYPCD = 999. It's stand age was recorded as 0, but stocking at 3... so I'm assuming that they're seedlings. The designated forest types seem like a reasonable comparison for the historical QQ data, with the exception of California black oak (n = 4), canyon live oak (n = 2), tanoak, (n = 1), and mountain hemlock (n = 1). After aggregating tree data, removed hardwoods from dataset and found that 6 plots had live BA < 9 m^2. After accounting for these constraints, our sample size would be 76 - 54 federally/state-owned and 22 privately/native-owned. Proceeded with aggregating tree data for each plot and exported a .csv file.

```{r}
#Create function to estimate density (trees/ha),
#Where x = TPA_UNADJ (TPA of a given tree),
#and 0.4047 is the conversion (ac to ha).
density.FUN <- function(x)
{density <- sum(x)/0.4047}

#Create function to estimate live BA (sq m/ha),
#Where a = DBH (in) of a given tree, b = TPA_UNADJ (TPA of that tree),
#0.00694444 is the conversion from sq in to sq ft,
#and 0.229568 is the conversion (sq ft/ac to sq m/ha).
liveba.FUN <- function(a,b)
{liveba <-  (sum(((pi*(a/2)^2)*0.00694444)*b))*0.229568}

#Import FIA condition and tree list data.
#Import forest and species reference lists from FIA.
FIA_TREE <- read.csv(here("large_files/CA_FIA", "CA_TREE.csv"))
FIA_COND <- read.csv(here("large_files/CA_FIA", "CA_COND.csv"))
FIA_COND <- FIA_COND[,-which(sapply(FIA_COND, function(x)all(is.na(x))))] #Delete empty columns
FIA_TREE <- FIA_TREE[,-which(sapply(FIA_TREE, function(x)all(is.na(x))))] #Delete empty columns
fia_forest_types <- read.csv(here("large_files/CA_FIA", "REF_FOREST_TYPE.csv"))
fia_species_list <- read.csv(here("large_files/CA_FIA", "REF_SPECIES.csv"))

#Renaming CN column to PLT_CN to join COND and TREE tables.
#According to FIA user guide,
#CN of plot data = PLT_CN of COND/TREE data (foreign key)
elevation <- elevation %>%
 rename(PLT_CN = CN)

#Joining COND table with subset of FIA data (PLT_CN is the foreign key).
#Extracting columns describing "conditions" (e.g. condition code, stand age, disturbance, stocking).
#Calculate number of conditions for each plot.
#Start with 182 plots.
#Remove plots with more than 1 condition (60 plots removed; 110 plots remain).
#Remove plots with fire damage (29 plots removed; 81 plots remain).
#Removed plots where FORTYPCD = 999 (1 plot with stand age = 0 (seedlings only), 80 plots remain).
#End with 80 plots.
condition_join <- inner_join(FIA_COND, elevation, by = c("PLT_CN", "INVYR")) %>%
  dplyr::select(PLT_CN,
                INVYR,
                ELEV_met,
                COND_STATUS_CD,
                STDAGE,
                FORTYPCD,
                OWNGRPCD,
                DSTRBCD1,
                CONDPROP_UNADJ,
                GSSTKCD,
                BALIVE,
                x.coord,
                y.coord) %>%
  group_by(PLT_CN) %>%
  mutate(number.conditions = length(COND_STATUS_CD)) %>%
  ungroup() %>%
  filter(number.conditions == 1) %>%
  filter(DSTRBCD1 < 30 | DSTRBCD1 > 32) %>%
  filter(FORTYPCD != 999)

#Creating a column FORTYPCD to join forest type data with condition data.
fia_forest_types <- fia_forest_types %>%
  mutate(FORTYPCD = VALUE)

#Joining forest type data with cond_check data.
#Start with 80 plots.
#Looked at forest types and some may not be good for comparison.
#Removed California black oak (n = 3), canyon live oak (n = 1), tanoak (n = 1)
#End with 76 plots.
forest_types_join<- inner_join(condition_join, fia_forest_types, by = "FORTYPCD") %>%
  filter(MEANING != "California black oak") %>%
  filter(MEANING != "Tanoak") %>%
  dplyr::select(PLT_CN,
                FORTYPCD,
                MEANING)

#Joining tree data with type_check data.
#Start with 76 plots.
#Subsetting relevant columns for analysis.
#Keeping only live trees with a DBH >= 12 inches (cut-off DBH for Collins Pine QQ data).
tree_join <- inner_join(forest_types_join, FIA_TREE, by = c("PLT_CN")) %>%
  dplyr::select(PLT_CN,
         STATUSCD,
         DIA,
         TPA_UNADJ,
         SPCD) %>%
  filter(STATUSCD == 1 & DIA >= 12)

#Joining species data with tree_join data.
#Start with 76 plots.
#Species include: ABCO, ABMA, ACMA (A. macrophyllum), ALRH (A. rhombifolia), CADE,
#PICO, PIJE, PILA, PIMO (p. monticola), PIPO, 
#POBAT (P. balsimifera spp. trichocarpa), PSME, QUCH, and QUKE.
#Removed hardwoods from dataset.
#None of the plots contained 100% hardwoods, 76 plots still remain.
#Calculate density of trees (>12 in DBH) for each plot.
#Calculate density of trees 12-24 DBH, 24-36 DBH, 36+ DBH.
#Calculate live BA in sq m/ha for each plot.
#Removed plots where live BA < 9 m^2/ha (5 plots removed; 71 plots remain).
#Estimating live BA by species for each plot.
#Collapsing rows to have single value for each species in each plot.
#Extracting relevant columns for analysis.
#Creating columns of live BA for each species.
#Collapsing rows to have single value for each species in each plot.
#Replacing NA with 0 for live BA by species.
#Renaming column headers to match Jens' code.
#Estimating pine ratio (BA pinus spp/total live basal area)
#End with 71 plots.
species_join <- inner_join(tree_join, fia_species_list, by = "SPCD") %>%
  filter(SPECIES_SYMBOL != "QUCH2") %>%
  filter(SPECIES_SYMBOL != "QUKE") %>%
  filter(SPECIES_SYMBOL != "ACMA3") %>%
  filter(SPECIES_SYMBOL != "ALRH2") %>%
  filter(SPECIES_SYMBOL != "POBAT") %>%
  group_by(PLT_CN) %>%
  mutate(DENS_12plus = density.FUN(TPA_UNADJ),
         DENS_12_24 = density.FUN(TPA_UNADJ[which(between(DIA, 11.99, 23.99))]),
         DENS_24_36 = density.FUN(TPA_UNADJ[which(between(DIA, 23.99, 35.99))]),
         DENS_36plus = density.FUN(TPA_UNADJ[which(DIA>35.99)]),
         BALIVE_sqmha = liveba.FUN(DIA, TPA_UNADJ)) %>%
  ungroup() %>%
  filter(BALIVE_sqmha > 8.9) %>%
  group_by(PLT_CN, SPECIES_SYMBOL) %>%
  mutate(BA_met = liveba.FUN(DIA, TPA_UNADJ)) %>%
  distinct(PLT_CN, .keep_all = TRUE) %>%
  dplyr::select(PLT_CN,
                DENS_12plus,
                DENS_12_24,
                DENS_24_36,
                DENS_36plus,
                BALIVE_sqmha,
                SPECIES_SYMBOL,
                BA_met) %>%
  ungroup() %>%
  spread(., "SPECIES_SYMBOL", "BA_met") %>%
  replace(is.na(.), 0) %>%
  rename(ABCO_met = ABCO,
         ABMA_met = ABMA,
         CADE_met = CADE27,
         PICO_met = PICO,
         PIJE_met = PIJE,
         PILA_met = PILA,
         PIMO_met = PIMO3,
         PIPO_met = PIPO,
         PSME_met = PSME) %>%
  group_by(PLT_CN) %>%
  mutate(pine_fraction = (PIJE_met + PILA_met + PIPO_met + PIMO_met)/BALIVE_sqmha) %>%
  ungroup()

#Joining tables to get all information for plots.
#Final FIA dataset has 71 plots.
d_FIA <- inner_join(species_join, condition_join, by = "PLT_CN")
d_FIA <- inner_join(d_FIA, forest_types_join, by = c("PLT_CN", "FORTYPCD"))

#Of those plots, 47 plots are under federal/state ownership,
#And 24 plots are under private ownership.
owner.check <- d_FIA %>%
  group_by(OWNGRPCD) %>%
  summarise(count = length(OWNGRPCD))

#Export extracted FIA data to .csv file
write.csv(d_FIA, here("Data/Derived", "d_FIA.csv"), row.names = FALSE)

#Converting final FIA dataset to .shp
d_FIA_shp <- d_FIA %>%
  st_as_sf(., coords = c("x.coord", "y.coord")) %>%
  st_set_crs(., st_crs(FIA.sf.clip))

st_write(d_FIA_shp, "GIS/d_FIA.shp")
```

Section III: Creating graphs of outputs

```{r}
#Chaning options in R to  3 significant figures.
options(digits = 3)

#Setting theme for figures
theme_set(theme_bw() + 
            theme(text = element_text(size = 12, family="serif"), 
        strip.background = element_blank()))

#Creating a column designating land ownership
#Codes < 40 are Federal- and State-owned
#Code = 40 is Private- an Native-owned
FIA_BA <- d_FIA %>%
  mutate(ownership = ifelse(OWNGRPCD < 40, "Federal/State",
                            ifelse(OWNGRPCD == 40, "Private/Native", NA)))

#Separating dataframes by ownership
Federal_BA <- FIA_BA %>%
  filter(ownership == "Federal/State")
Private_BA <- FIA_BA %>%
  filter(ownership == "Private/Native")

#Creating a color palette for histogram
palette.1 <- c(rep("firebrick3", 5), 
          rep("sienna2", 5), 
          rep("tan1",5),
          rep("khaki1", 5), 
          rep("lightskyblue1", 5),
          rep("steelblue", 5))

###########################################################################
#Live BA
###########################################################################
#Histogram of live BA
ggplot(FIA_BA, aes(BALIVE_sqmha)) + 
  geom_histogram(color = "black",
                 fill = palette.1) + 
  xlab(bquote("Live BA"~(m^2~ha^-1))) +
  ylab("Count") + 
  ggsave(here("Figures/Lassen-Plumas", "BALive.png"),
         width = 6, height = 4, dpi = 500)

#Histogram of live BA by ownership
#Darker is private, lighter shade is federal
ggplot(Private_BA, aes(BALIVE_sqmha)) + 
  geom_histogram(color = "black",
                 fill = palette.1) + 
  geom_histogram(data = Federal_BA, aes(BALIVE_sqmha),
                 color = "black",
                 fill = palette.1,
                 alpha = 0.5,
                 linetype = "dashed") + 
  xlab(bquote("Live BA"~(m^2~ha^-1))) +
  ylab("Count") + 
  ggsave(here("Figures/Lassen-Plumas", "BALive_ownership.png"),
         width = 6, height = 4, dpi = 500)

#Density plot of live BA by ownership
ggplot(FIA_BA, aes(BALIVE_sqmha, fill = ownership)) + 
  geom_density(alpha = 0.5) + 
  scale_fill_manual(values = c("goldenrod2",
                               "dodgerblue3")) + 
  xlab(bquote("Live BA"~(m^2~ha^-1))) +
  coord_cartesian(expand = FALSE) + 
  ylab("Density") +
  theme(legend.title = element_blank()) +
  ggsave(here("Figures/Lassen-Plumas", "BALive_density_ownership.png"),
         width = 6, height = 4, dpi = 500)

##############################################################################
#Pine fraction
##############################################################################
#Histogram of Pine fraction
ggplot(FIA_BA, aes(pine_fraction)) + 
  geom_histogram(color = "black",
                 fill = palette.1) + 
  xlab("Pine fraction") +
  ylab("Count") + 
  ggsave(here("Figures/Lassen-Plumas", "Pine_fraction.png"),
         width = 6, height = 4, dpi = 500)  
  
#Histogram of Pine fraction by ownership
#Darker is private, lighter shade is federal
ggplot(Private_BA, aes(pine_fraction)) + 
  geom_histogram(color = "black",
                 fill = palette.1) + 
  geom_histogram(data = Federal_BA, aes(pine_fraction),
                 color = "black",
                 fill = palette.1,
                 alpha = 0.5,
                 linetype = "dashed") + 
  xlab("Pine fraction") +
  ylab("Count") + 
  ggsave(here("Figures/Lassen-Plumas", "Pine_ownership.png"),
         width = 6, height = 4, dpi = 500)

#Density plot of Pine fraction by ownership
ggplot(FIA_BA, aes(pine_fraction, fill = ownership)) + 
  geom_density(alpha = 0.5) + 
  scale_fill_manual(values = c("goldenrod2",
                               "dodgerblue3")) + 
  xlab("Pine fraction") +
  coord_cartesian(expand = FALSE) + 
  ylab("Density") +
  theme(legend.title = element_blank()) +
  ggsave(here("Figures/Lassen-Plumas", "Pine_density_ownership.png"),
         width = 6, height = 4, dpi = 500)

#############################################################################
#TPH
#############################################################################
#Creating a datframe to graph TPH
#Gathering columns (long format) to designate diameter classes
#Removing entries where TPH = 0
FIA_density <- FIA_BA %>%
  group_by(PLT_CN) %>%
  gather(., "DBH.class", "TPH", 2:5) %>%
  filter(TPH != 0) %>%
  ungroup() %>%
  mutate(dbh = ifelse(DBH.class == "DENS_12plus", "All Trees",
                      ifelse(DBH.class == "DENS_12_24", "12-24in",
                             ifelse(DBH.class == "DENS_24_36", "24-36in",
                                    ifelse(DBH.class == "DENS_36plus", ">= 36in", NA)))))

#Reordering diameter classes for graph
FIA_density$dbh <- factor(FIA_density$dbh, levels = c("All Trees",
                                                      "12-24in",
                                                      "24-36in",
                                                      ">= 36in"))

#Separating dataframes by ownership
Federal_density <- FIA_density %>%
  filter(ownership == "Federal/State")
Private_density <- FIA_density %>%
  filter(ownership == "Private/Native")

#Establishing a color palette for facet histograms
palette.2 <- rep(c(rep("firebrick3", 5), 
          rep("sienna2", 5), 
          rep("tan1",5),
          rep("khaki1", 5), 
          rep("lightskyblue1", 5),
          rep("steelblue", 5)), 4)

#Histogram of TPH by diameter class
ggplot(FIA_density, aes(TPH)) +
  geom_histogram(color = "black",
                 fill = palette.2,
                 origin = -0.5) + 
  facet_wrap(~dbh, scales = "free") + 
  ylab("Count") + 
  ggsave(here("Figures/Lassen-Plumas", "TPH.png"),
         width = 6, height = 4, dpi = 500)

#Histogram of TPH by diameter class AND ownership
#Dark is private, lighter shade is federal
ggplot(Private_density, aes(TPH)) +
  geom_histogram(color = "black",
                 fill = palette.2,
                 origin = -0.5) + 
  facet_wrap(~dbh, scales = "free") + 
  geom_histogram(data = Federal_density, aes(TPH),
                 color = "black",
                 fill = palette.2,
                 alpha = 0.5,
                 origin = -0.5,
                 linetype = "dashed") + 
  ylab("Count") + 
  ggsave(here("Figures/Lassen-Plumas", "TPH_ownership.png"),
         width = 6, height = 4, dpi = 500)

#Density plot of TPH by diameter class AND ownership
ggplot(FIA_density, aes(TPH, fill = ownership)) + 
  geom_density(alpha = 0.5) + 
  scale_fill_manual(values = c("goldenrod2",
                               "dodgerblue3")) + 
  xlab("TPH") +
  coord_cartesian(expand = FALSE) + 
  ylab("Density") +
  facet_wrap(~dbh, scales = "free") +
  theme(legend.title = element_blank()) +
  ggsave(here("Figures/Lassen-Plumas", "TPH_density_ownership.png"),
         width = 6, height = 4, dpi = 500)
```

